Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-25 paper/paper.tex
Typo fixes

The container encoding \coq{ITreeF} has two shapes, one for leaves and one for
nodes, and nodes have two positions, one for each sublist \coq{l r : X} in
\coq{i_node}.
\begin{coqcode}
Inductive Tshape L A := | Leaf | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
The container for \coq{ITreeF} is denoted by \coq{TreeF}. For convenience, we
define the wrappers \coq{a_leaf} and \coq{a_node} to construct values of the
extension of this container.
Now, the container for quicksort is given by \coq{TreeF int}.
In the definition below, we use \coq{posL} and \coq{posR} as wrappers for
\coq{Lbranch} and \coq{Rbranch} with validity proofs.
\begin{coqcode}
Definition qsplit_f (x : list int) : App (TreeF int) (list int) :=
  match x with
  | nil => a_leaf tt
  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
  end.

Definition merge_f (x : App (TreeF int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ _ => fun _ => nil
  | Node _ h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
\end{coqcode}
The proofs that these are proper morphisms (called \coq{merge} and \coq{qsplit})
are automatically discharged by our tactics, but we still have to prove that the
coalgebra \coq{qsplit} is recursive. In this specific case we have to prove that
the two sublists are smaller than the input list. In general, we are required to
prove that the input is related to the data inside the structure functor, that
is that the element of the carrier of the coalgebra is related to the content of
the functor by some well-founded relation. Here we give a specialised property, the
full general property can be found in the formalisation:
\begin{coqcode}
Definition respects_relation `{setoid A} (c : Coalg F B) B} (R : B -> B -> Prop)
:= forall x (p : Pos (shape (c x))), R (length (cont (c x) p)) (length x).
\end{coqcode}
The resulting code extracted from \coq{hylo merge qsplit} is similar to a
hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
The correctness proof is done by proving the following statement:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
\end{coqcode}
that is any list produced by quicksort is sorted and is some permutation of the original one.

@@

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.
\newcommand{\Id}{\text{Id}}
\subsection{Knapsack}\label{sec:knap}
Dynamorphisms are hylomorphisms where the algebra has access to the memory table
of the result of all previous recursive calls. To implement this we need to
formalise the idea of a memory table which is in our case simply represented by
the cofree comonad. For a functor $G : \CatC \to \CatC$ the cofree comonad is
defined as the greatest solution to the following equation
\[
  G_{\infty} A \iso A \times G (G_{\infty} A)
\]
In words, these are the (possibly) infinite trees which branch out with shape
$G$. For example, for the identity functor $\Id_{\infty} A$ is the type of
infinite streams. For $G X = 1 + X$, the type $G_{\infty} A$ is isomorphic to
$i \times (1 + G_{\infty} A)$. This is the type representing our memory table
with the operation $\head_{A} : G_{\infty} A$ being the result of the previous
recursive call. Notice that in the case of a recursive $G$-coalgebra this type
only contains finite $G$-trees and so we can view $G_{\infty}A$ as the finite
$G$-trees thus being able to inspect all the previous recursive calls. This has
been expounded thoroughly by Hinze et al.~\cite{HinzeWG15}.

>>>>

The container encoding \coq{ITreeF} has two shapes, one for leaves and one for
nodes, and nodes have two positions, one for each sublist \coq{l r : X} in
\coq{i_node}.
\begin{coqcode}
Inductive Tshape A := | Leaf | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
The container for \coq{ITreeF} is denoted by \coq{TreeF}. For convenience, we
define the wrappers \coq{a_leaf} and \coq{a_node} to construct values of the
extension of this container.
Now, the container for quicksort is given by \coq{TreeF int}.
In the definition below, we use \coq{posL} and \coq{posR} as wrappers for
\coq{Lbranch} and \coq{Rbranch} with validity proofs.
\begin{coqcode}
Definition qsplit_f (x : list int) : App (TreeF int) (list int) :=
  match x with
  | nil => a_leaf
  | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in a_node h l r
  end.

Definition merge_f (x : App (TreeF int) (list int)) : list int :=
  let (sx, kx) := x in
  match sx return (Container.Pos sx -> _) -> _ with
  | Leaf _ => fun _ => nil
  | Node h => fun k => List.app (k (posL h)) (h :: k (posR h))
  end kx.
\end{coqcode}
The proofs that these are proper morphisms (called \coq{merge} and \coq{qsplit})
are automatically discharged by our tactics, but we still have to prove that the
coalgebra \coq{qsplit} is recursive. In this specific case we have to prove that
the two sublists are smaller than the input list. In general, we are required to
prove that the input is related to the data inside the structure functor, that
is that the element of the carrier of the coalgebra is related to the content of
the functor by some well-founded relation. Here we give a specialised property, the
full general property can be found in the formalisation:
\begin{coqcode}
Definition respects_relation `{setoid A} (c : Coalg F B) B} (R : B -> B -> Prop)
:= forall x (p : Pos (shape (c x))), R (length (cont (c x) p)) (length x).
\end{coqcode}
The resulting extracted code is similar to a hand-written implementation:
\begin{ocamlcode}
let rec qsort = function
  | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
    let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
    app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
The correctness proof is done by proving the following statement:
\begin{coqcode}
forall (l : list int), Sorted (hylo merge qsplit l) /\ Perm l (hylo merge qsplit l).
\end{coqcode}
that is any list produced by quicksort is sorted and is some permutation of the original one.

@@

%% Our formalisation contains a proof that \coq{natural} is indeed a natural
%% transformation, which relies on the fact that it preserves the structure of the
%% shapes and, therefore, the validity of the positions.
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
The extracted OCaml code is a single recursive traversal. Note the similarity to
a hand-written implementation, and that it has been derived by fusing two
different recursive programs using regular Coq rewrite tactics.
\newcommand{\Id}{\text{Id}}
\subsection{Knapsack}\label{sec:knap}
Dynamorphisms are hylomorphisms where the algebra has access to the memory table
of the result of all previous recursive calls. To implement this we need to
formalise the idea of a memory table which is in our case simply represented by
the cofree comonad. For a functor $G : \CatC \to \CatC$ the cofree comonad is
defined as the greatest solution to the following equation
\[
  G_{\infty} A \iso A \times G (G_{\infty} A)
\]
In words, these are the (possibly) infinite trees which branch out with shape
$G$. For example, for the identity functor $\Id_{\infty} A$ is the type of
infinite streams. For $G X = 1 + X$, the type $G_{\infty} A$ is isomorphic to
$A \times (1 + G_{\infty} A)$. This is the type representing our memory table
with the operation $\head_{A} : G_{\infty} A$ being the result of the previous
recursive call. Notice that in the case of a recursive $G$-coalgebra this type
only contains finite $G$-trees and so we can view $G_{\infty}A$ as the finite
$G$-trees thus being able to inspect all the previous recursive calls. This has
been expounded thoroughly by Hinze et al.~\cite{HinzeWG15}.
